---
title: "Exercises"
author: "Adrien Osakwe"
format: html
editor: visual
---

### Exercise 1

Poisson model has mass function

$$
f_Y(y;\theta) = \frac{\theta^yexp\{-\theta\}}{y!}
$$

for $y \in \mathbb{Z}^+$

Explore the posterior density $\pi_n(\theta)$ using two priors

-   $Gamma(\alpha_0,\beta_0)$

-   prior determined by the assumption that $\phi = log\theta \sim Normal(\eta_0,\tau_0^2)$ a priori. In other words, that the prior is a **log-normal distribution**

We have the following observations

|       |     |     |     |     |     |     |     |
|-------|-----|-----|-----|-----|-----|-----|-----|
| y     | 0   | 1   | 2   | 3   | 4   | 5   | 6   |
| Count | 2   | 6   | 7   | 16  | 11  | 6   | 2   |

Assuming that the likelihood is independent **given** $\theta$ (de Finetti), multiplying the joint likelihood by the Gamma prior gives us a Gamma posterior with observation-dependent parameters (due to gamma-poisson conjugacy - see [Conjugate Distributions](./Conjugate_Dist.qmd)).

$$
\pi_n(\theta) \sim Gamma(\alpha_0 + s_n,\beta_0 + n) \\
s_n = \sum^n_{i=1}y_i
$$

We can now plot the posterior over all possible $\theta \in \Theta$.

```{r}
a <- 1
b <- 1
y <- c(rep(0,2),rep(1,6),rep(2,7),rep(3,16),rep(4,11),rep(5,6),rep(6,2))
s_n <- sum(y)
n <- length(y)

theta <- seq(0,5,by = 0.01)
ll <- sapply(theta,function(thet){
  prod(dpois(y,thet))
})
#scalling likelihood to be visible in plot
M <- max(dgamma(theta,a+s_n,b+n))/max(ll)
plot(theta,dgamma(theta,a+s_n,b+n),type = 'l',col = 'red',lwd = 2,
     xlab = expression(theta),
     ylab = 'Density')
lines(theta,dgamma(theta,a,b),col = 'blue',lwd = 2)
lines(theta,ll*M,col = 'orange',lwd = 2)
legend('topright',legend = c('Posterior',' Gamma Prior','Scaled Joint-Likelihood'),col = c('red','blue','orange'),lwd = 2)
```

```{r}
eta <- 0
tau <- 10
#Can treat the joint-likelihood as a gamma dist with a = sn and b = n to avoid need for individual observations
lnorm_pois <- function(eta,tau,theta,sn,n){
  calc <- dgamma(theta,sn,n)*dnorm(log(theta),eta,tau)
  return(calc)
}
norm_const <- integrate(lnorm_pois,eta = eta,tau =tau,sn = s_n,n = n,lower = 0, upper = 10)

M <- max(lnorm_pois(eta,tau,theta,s_n,n)/norm_const$value)/max(ll)
S <- max(lnorm_pois(eta,tau,theta,s_n,n))/max(dnorm(log(theta),eta,tau))
plot(theta,lnorm_pois(eta,tau,theta,s_n,n)/norm_const$value,type = 'l',col = 'red',lwd = 2,
     xlab = expression(theta),
     ylab = 'Density')
lines(theta,dnorm(log(theta),eta,tau)*S,col = 'blue',lwd = 2)
lines(theta,ll*M,col = 'orange',lwd = 2)
legend('topright',legend = c('Posterior',' Log-Norm Prior','Scaled Joint-Likelihood'),col = c('red','blue','orange'),lwd = 2)
```

### Exercise 2
